Gaia's constraints on the $\sim$$150\,M_\odot$ runaway star candidate VFTS682

### 2018-06-08 VFTS682_APJL.tex
editing

 How very massive stars form is still among the most intriguing open questions in stellar astrophysics.  VFTS 682 is among the most extreme massive stars known, with an inferred initial mass of about $150\,M_\odot$. It is located in the Tarantula nebula at a projected distance of 29\,pc from the central cluster R136.  The absence of other massive stars in its immediate vicinity lead to two intriguing possible hypotheses posed in earlier work. Either it formed in relative isolation through a new mode of star formation or it was ejected dynamically from the central cluster.

@@

If future data confirms the runaway nature, this would make the VFTS682 the most massive runaway star known to date. 
While we cannot prove this solidly from the current data, we do consider this hypothesis plausible. The central cluster is known to harbor several other ``monster stars'' similar spectral types and high inferred masses.  Moreover, the current record holders are O type stars including VFTS 16 which provides direct evidence that R136 is indeed capable of ejecting some of its most massive members.  We dicuss further implications and caveats.

\end{abstract}

@@

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to  form preferentially in the dense cores. In this picture, field stars are primarily the result of the dissolution of dense groups.  However, a significant population of massive stars exists in relative isolation, far from dense clusters or OB associations and their origin remains matter of debate \citep{gvaramadze:12, lamb:16,ward:18}.   One hypothesis to explain the population of relatively isolated massive stars is that they formed in the field. The alternative hypothesis is that these massive stars were ejected from the clusters in which they formed.  Such ejections may result from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems at the death of the companion  star \citep[e.g.,][]{zwicky:57, blaauw:61, renzo:18}.

@@

One of the most extreme examples that has been considered in this debate is the very massive star VFTS682  \cite{bestenlehner:11, bressert:12}. This star is located in the field of the 30 Doradus region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey \citep[VFTS,][]{evans:11}. It is a hydrogen-rich Wolf-Rayet star of spectral type is WNh5. Spectral analysis and comparison with evolutionary models lead to an inferred present-day mass of $140^{+30}_{-16}\,M_\odot$ corresponding to an initial mass of $150.0^{+30}_{-17}\,M_\odot$
\citep{schneider:18}. 
%
%$137.8^{+27.5}_{-15.9}\,M_\odot$,
%
This makes VFTS682 one of the most massive stars known and one of the most extreme objects in the region.
%
From the spectral point of view, it is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
 in the core of the R136 cluster \citep{dekoter:97,crowther:10, crowther:16}. 
 %
 In particular, a remarkable similarity exist between the
spectrum of VFTS682 and R136a3 for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. R136 hosts
at least two more very massive WN5h stars, R136a1 and R136a2, whose
estimated current masses are even higher.

@@

VFTS682 stands out by its relative isolation at a projected distance of $\sim$$29$\,pc from  the star cluster R136.
\citet{bestenlehner:11} discuss the possible explanation for the offset. They consider formation in situ as an isolated massive star, but remark that this poses a challenge for starformation theories. The alternative, ejection from  R136, appears to be reasonable. N-body simulations of dynamical interaction in star clusters indicate that the ejection of very massive stars like VFTS682  is expected \citep[e.g.][]{fujii:11, banerjee:12}.  This is supported by the recent findings of other massive runaway stars in the region based on proper motion studies.   \citet{platais:15,platais:18} analyze multi-epoch \emph{Hubble Space Telescope} (HST) photometry and identify 10 stars, which appear to be ejected from R136.   \citet{lennon:18} investigate the kinematics of  isolated O-type stars in the region using the second \emph{Gaia} data release \cite[DR2,][]{gaia:16,brown:18} and show that the proper motion, postion and direction of the $\sim100\Msun$ star VFTS16 is consistent with a runaway origin from R136.

@@

In this paper we present an analysis of the new kinematical constraints for VFTS682 provided by \emph{Gaia} DR2 and constraints from HST proper data  by \citet{platais:18}.   We discuss the implications for the hypothesis that VFTS682 is a slow runaway star ejected from R136 and discuss the caveats and implications.

% which was not part of the sample of \citet{lennon:18} because of its  spectral type.

@@

\cite{bressert:12} note that VFTS682 shows a significant radial velocity offset compared to the nebular lines from the gas surrounding it, suggesting peculiar motion in the line of sight direction.  \citet[ ][]{bestenlehner:11} quote a radial velocity measured from the NV $\lambda4944$ of   $300\pm10\kms$. The average radial velocity of the 30 Doradus region is 
$270\pm10\kms$. Whether this measurement truly reflects a peculiar radial velocity is debated. The determination of radial velocities for emission-lines stars is difficult because of the presence of an optically thick, non-homogeneous, and variable wind obscuring the photosphere. The radial velocity errors quoted above are only statistical, and do not quantify the (possibly larger)
systematic errors.  \SdM{TODO: Possibly mention and cite paper in prep by Paco and student on Xshooter data?  Possibly mention (or even use) on their measurement for the RV?}

%We implicitly assume that the use of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not introduce significant errors.

% observations; the first epoch (GO-12499; P.I.: D.~J.~Lennon)
% in October 3-8 2011 and the second epoch (GO-13359) in
% October 6-11 2014, both using the F775W filter.
% The second epoch was designed to match the pointing and
% orientation of the first epoch as exactly as possible so
% that the displacement measurements could be purely differential.

\subsection{HST astrometry for VFTS682}
For large distances ($\gtrsim50$\,kpc), HST astrometry can compete or even exceed \emph{Gaia} DR2 data in quality.

Recently, \citet{platais:18} presented HST astrometry and proper motions in the 30 Doradus region (GO-12499; P.I.: D.~J.~Lennon). The brightest stars (V$<$14) are missing from the catalogue because of saturation, but the high extinction towards VFTS682 makes is fain enough to be included in this catalogue. The  proper motions are
listed in \Tabref{tab:vfts682}.

The catalog from \citet{platais:18} lists the proper motion
components of VFTS682 relative to the field, providing a
measure of the motion of this star completely independent from
ours. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$, which added in quadrature give a relative proper motion of
$\delta \mu^{HST} =0.20 \pm 0.15\,\mathrm{mas\ yr^{-1}}$.
% \begin{equation}
%   \label{eq:pm_around_HST}
%   \delta \mu^{HST} = \sqrt{\left(\delta\mu_\mathrm{RA}^{HST}\right)^2+\left(\delta\mu_\mathrm{DEC}^{HST}\right)^2}
%   = 0.20 \pm 0.15\,\mathrm{mas\
%   yr^{-1}} \ \ .
% \end{equation}

These proper motion components can also be converted into projected
velocities assuming a distance of 50\,kpc, obtaining $\delta
v_\mathrm{RA}^{HST}=2\pm31\kms$ and $\delta
v_\mathrm{RA}^{HST}=47\pm24\kms$. % , corresponding to a two-dimensional
% projected velocity of $47\pm24\kms$.
Adding them in quadrature to the radial velocity
measured by \citet{bestenlehner:11} we obtain a three-dimensional velocity of
$v_\mathrm{pec}^{HST}=56 \pm 23  \kms$.
% \begin{equation}
%   \label{eq:speed_around_HST}
%   v_\mathrm{pec}^{HST} = \sqrt{\left(\delta v_\mathrm{RA}^{HST}\right)^2
%     +\left(\delta v_\mathrm{DEC}^{HST}\right)^2+\left(\delta
%       v_\mathrm{rad}^{VFTS}\right)^2} = 56 \pm 23
%   \kms \ .
% \end{equation}

@@

% We calculate the averaged proper motion components for the whole
% region using 
% \begin{equation}
%   \label{eq:mean}
%   \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
%   \mu_i^2}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i^2}} \ \ , \
%   \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
%   \frac{1}{\Delta \mu_i^2}} \ \ ,
% \end{equation}
% where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
% on the proper motion component reported by \emph{Gaia}. The sums run over
% all the $N=66$ stars in our selected sample. We evaluate each proper motion
% component separately.
For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error on
the distance determination ($\lesssim2\%$) when converting proper motions into
physical velocities. % The data retrieved, and the ipython notebook used for the analysis
% presented here will be made available at \todo{git repo bitbucket?}.

>>>>

 How very massive stars form is still among the most intriguing open
 questions in stellar astrophysics.  VFTS 682 is among the most
 extreme massive stars known, with an inferred initial mass of about
 $150\,M_\odot$. It is located in 30 Doradus at a projected distance of 29\,pc from the central cluster R136.  The absence of other massive stars in its immediate vicinity lead to two intriguing possible hypotheses posed in earlier work. Either it formed in relative isolation through a new mode of star formation or it was ejected dynamically from the central cluster.

@@

If future data confirms the runaway nature, this would make the VFTS682 the most massive runaway star known to date. 
While we cannot prove this solidly from the current data, we do
consider this hypothesis plausible. The central cluster is known to
harbor several other stars more massive than $150\,M_\odot$ similar in
spectral type.  Moreover, the current record holders are O type stars
including VFTS 16 which provides direct evidence that R136 is indeed
capable of ejecting some of its most massive members. 
\end{abstract}

@@

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to reside in the innermost cores. In this picture, field stars are primarily the result of the dissolution of dense groups.  However, a significant population of massive stars exists in relative isolation, far from dense clusters or OB associations and their origin remains matter of debate \citep{gvaramadze:12, lamb:16,ward:18}.   One hypothesis to explain the population of relatively isolated massive stars is that they formed in the field. The alternative hypothesis is that these massive stars were ejected from the clusters in which they formed.  Such ejections may result from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems at the death of the companion  star \citep[e.g.,][]{zwicky:57, blaauw:61, renzo:18}.

@@

One of the most extreme examples that has been considered in this debate is the very massive star VFTS682  \citep[][]{bestenlehner:11, bressert:12}. This star is located in the field of the 30 Doradus region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey \citep[VFTS,][]{evans:11}. It is a hydrogen-rich Wolf-Rayet star of spectral type WNh5. Spectral analysis and comparison with evolutionary models lead to an inferred present-day mass of $\sim$$140^{+30}_{-16}\,M_\odot$ corresponding to an initial mass of $\sim$$150^{+30}_{-17}\,M_\odot$
\citep{schneider:18}. 
%
%$137.8^{+27.5}_{-15.9}\,M_\odot$,
%
This makes VFTS682 one of the most massive stars known and one of the most extreme objects in the region.
%
From the spectral point of view, it is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
 in the core of the R136 cluster \citep{dekoter:97,crowther:10, crowther:16}. 
 %
 In particular, a remarkable similarity exist between the
spectrum of VFTS682 and R136a3 for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. R136 hosts
at least two more very massive WN5h stars, R136a1 and R136a2, whose
estimated current masses are even higher.

@@

VFTS682 stands out by its relative isolation at a projected distance of 119.4 arcseconds, corresponding to 
$\sim$$29$\,pc, from  the star cluster R136. \citet{bestenlehner:11}
considered two possible explanation for the offset: either
the star formed in situ as an isolated massive star, % , but remark that this
                                % poses a challenge for starformation
                                % theories (They don't)
or it was ejected from  R136. N-body simulations % of dynamical interaction in star clusters
indicate that the ejection of very massive stars like VFTS682 is expected \citep[e.g.][]{fujii:11, banerjee:12}.  This is supported by the recent findings of other massive runaway stars in the region based on proper motion studies.   \citet{platais:15,platais:18} analyze multi-epoch \emph{Hubble Space Telescope} (HST) photometry and identify 10 stars, which appear to be ejected from R136.   \citet{lennon:18} investigate the kinematics of  isolated O-type stars in the region using the second \emph{Gaia} data release \cite[DR2,][]{gaia:16,brown:18} and show that the proper motion, postion and direction of the $\sim100\Msun$ star VFTS16 is consistent with a runaway origin from R136.

@@

In this paper we present an analysis of the new kinematical
constraints for VFTS682 provided by \emph{Gaia} DR2 and constraints
from HST proper data  by \citet{platais:18}.   We discuss the
implications for the hypothesis that VFTS682 is a slow runaway star
ejected from R136.
% which was not part of the sample of \citet{lennon:18} because of its  spectral type.

@@

\cite{bressert:12} note that VFTS682 shows a significant radial
velocity offset compared to the nebular lines from the gas surrounding
it, suggesting peculiar motion in the line of sight direction.
\citet{bestenlehner:11} quote a radial velocity measured from the
N{\footnotesize V} $\lambda4944$ of   $300\pm10\kms$. The average radial velocity of the 30 Doradus region is 
$270\pm10\kms$. Whether this measurement truly reflects a peculiar radial velocity is debated. The determination of radial velocities for emission-lines stars is difficult because of the presence of an optically thick, non-homogeneous, and variable wind obscuring the photosphere. The radial velocity errors quoted above are only statistical, and do not quantify the (possibly larger)
systematic errors.  \todo{Possibly mention and cite paper in prep by Paco and student on Xshooter data?  Possibly mention (or even use) on their measurement for the RV?}

For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error on
the distance determination ($\lesssim2\%$) when converting proper motions into
physical velocities.
%We implicitly assume that the use of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not introduce significant errors.

\subsection{HST astrometry for VFTS682}
% observations; the first epoch (GO-12499; P.I.: D.~J.~Lennon)
% in October 3-8 2011 and the second epoch (GO-13359) in
% October 6-11 2014, both using the F775W filter.
% The second epoch was designed to match the pointing and
% orientation of the first epoch as exactly as possible so
% that the displacement measurements could be purely differential.

For large distances ($\gtrsim50$\,kpc), HST astrometry can compete or even exceed \emph{Gaia} DR2 data in quality.

Recently, \citet{platais:18} presented HST astrometry and proper
motions in the 30 Doradus region (GO-12499; P.I.: D.~J.~Lennon). The
brightest stars (V$<$14) are missing from the catalogue because of
saturation, but the high extinction towards VFTS682 makes is faint
enough to be included. % in this catalogue.
The  proper motions are
listed in \Tabref{tab:vfts682}.

The catalog from \citet{platais:18} lists the proper motion
components of VFTS682 relative to the field, providing a
measure of the motion of this star completely independent from
\emph{Gaia}. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$, which added in quadrature give a relative proper motion of
$\delta \mu^{HST} =0.20 \pm 0.15\,\mathrm{mas\ yr^{-1}}$.
% \begin{equation}
%   \label{eq:pm_around_HST}
%   \delta \mu^{HST} = \sqrt{\left(\delta\mu_\mathrm{RA}^{HST}\right)^2+\left(\delta\mu_\mathrm{DEC}^{HST}\right)^2}
%   = 0.20 \pm 0.15\,\mathrm{mas\
%   yr^{-1}} \ \ .
% \end{equation}

These proper motion components can also be converted into projected
velocities assuming a distance of 50\,kpc, obtaining $\delta
v_\mathrm{RA}^{HST}=2\pm31\kms$ and $\delta
v_\mathrm{RA}^{HST}=47\pm24\kms$% , corresponding to a two-dimensional
% projected velocity of $47\pm24\kms$
.
Adding them in quadrature to the radial velocity
measured by \citet{bestenlehner:11} we obtain a three-dimensional velocity of
$v_\mathrm{pec}^{HST}=56 \pm 23  \kms$.
% \begin{equation}
%   \label{eq:speed_around_HST}
%   v_\mathrm{pec}^{HST} = \sqrt{\left(\delta v_\mathrm{RA}^{HST}\right)^2
%     +\left(\delta v_\mathrm{DEC}^{HST}\right)^2+\left(\delta
%       v_\mathrm{rad}^{VFTS}\right)^2} = 56 \pm 23
%   \kms \ .
% \end{equation}

@@

% We calculate the averaged proper motion components for the whole
% region using 
% \begin{equation}
%   \label{eq:mean}
%   \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
%   \mu_i^2}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i^2}} \ \ , \
%   \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
%   \frac{1}{\Delta \mu_i^2}} \ \ ,
% \end{equation}
% where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
% on the proper motion component reported by \emph{Gaia}. The sums run over
% all the $N=66$ stars in our selected sample. We evaluate each proper motion
% component separately.
